Accreditation: Post Shadbolt

### 2019-10-04 CEP20/CEP_accreditationpostshadbolt.tex
Draft of response to Shadbolt

@@

\author{Alastair irons}
\affiliation{%
  \institution{ Sunderland University}
  \city{Sunderland}
  \country{UK}
}
\email{alastair.irons@sunderland.ac.uk}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

@@

\section {What is it?}
A short description of the practice you're presenting

\section {Why are you doing it?}	What happened before? What is it changing / replacing / improving? What gap is it filling?

\section {Where does it fit?	}A short description of your teaching context. You may, for instance, include a description of intake, class size, curriculum sequence; anything that's necessary for others to understand your situation. How do things work at your institution?

\section {Does it work?}	How do you know? Give some evidence of effectiveness in context.
\section {Who else has done this?}	Where did you get the idea from? (If from published reports, please include references). How did you find out about it? Was it easy/hard to adopt? What did you change?
\section {What will you do next?}	Will you vary this, or develop it further?
\section{Why are you telling us this?}	What is interesting, or useful, about this to someone else?

>>>>

\usepackage{dirtytalk}

@@

\author{Alastair Irons}
\affiliation{%
  \institution{ Sunderland University}
  \city{Sunderland}
  \country{UK}
}
\email{alastair.irons@sunderland.ac.uk}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The promotion of quality via Accreditation by Professional, Statutory and Regulatory Bodies (PSRBs) is very much a feature of higher education in the united kingdom (UK).  In this changing  environment, there is a need for accreditation regimes to evolve in order to maximise the value they provide to Higher Education Institutions(HEIs).  This paper provides a summary in the context of one professional body (BCS, The Chartered Institute For IT) of 
\begin{enumerate}
\item what has happened in response to the recommendations of the Shadbolt review \cite{shadbolt2016shadbolt}
\item ongoing enhancement projects and the related future conversations that will help steer further enhancements
\end{enumerate}
\end{abstract}

@@

\section {What is it?}
%A short description of the practice you're presenting

The Shadbolt Review \cite{shadbolt2016shadbolt} investigated high unemployment rates graduates of computer science and related degrees in the United Kingdom and the role of degree accreditation in promoting employability.The Shadbolt review noted that the rates of graduate unemployment were declining and varied considerably between geographic location and type of HEI. The discipline was already acting proactively to address the issue.  Related professional bodies have enhanced their accreditation processes in response these challenges.

BCS has made significant adjustments to their accreditation processes since 2015. These were partly in response to the recommendations of Shadbolt but also to other changes in the sector.  These have been communicated to the assessor community and to those responsible for leading the development of HEI accreditation applications. However, there are many academics who are less intimately involved in accreditation, this paper serves to communicate these and future enhancements to them and the wider computer science education community.

\section {Why are you doing it?}	
%What happened before? What is it changing / replacing / improving? What gap is it filling?

The Shadbolt review \cite[p.~8]{shadbolt2016shadbolt} makes two main recommendations for the enhancement of accreditation regimes.
\begin{center} 
	\begin{textit} 
	"Recommendation 9 - Academic Accreditation of Degree Courses.

		BCS, IET and Tech Partnership should ensure that existing systems of degree course
	accreditation are flexible, agile, and enable HE providers to respond to changing
	demand and emerging technological trends and developments. Accreditation of courses
	should be focussed on outputs. Accrediting bodies should work to increase awareness
	and value of accreditation so that it is valued by HE providers, students and employers,
	and consider how their role can provide a forum for engagement between HE and
	employers..

	Recommendation 10 - Engaging industry in accreditation.

	Employers, through employer groups, such as Tech Partnership, should engage more
	consistently with HE providers and BCS \& IET to ensure accreditation is effective and
	reflects current industry demand.
	"
	\end{textit}
\end{center}

Criticisms of the accreditation of degree programmes is not new, there is a history of claiming the processes are unnecessarily bureaucratic and constrain  innovation \cite{Harvey2004}. Equally its value has been highlighted particularly in the context of a potentially globally mobile workforce, and the dangers of accreditation streams being revenues streams in their own right rather than for the benefit of a discipline or wider society \cite{Knight_2015}.

Hence, the challenge set for BCS and other accreditation providers that operate in the Computer Science discipline was broadly to maintain internationally recognised standards; reduce the perceived bureaucracy involved; respond to emerging technology trends and developments; promote enhancement and innovation; increase awareness and value of accreditation; and engage industry.

Defining quality and what is required for quality improvement has been discussed by a plethora of authors. One early author, Joseph Juran indicated as part of his Quality Control Handbook that a trilogy was required, namely quality planning, quality control and quality improvement \cite{juran1962quality}. This serves to indicate that quality improvement efforts should be informed by the data available and are never complete. As such there is a need to continue to enhance accreditation processes, as there are all quality review and improvement processes.

BCS is addressing these issues in the following ways.

\subsection{The value of accreditation}
What is the value of an accredited degree? This is a question that the BCS is in the process of exploring in much depth with a variety of stakeholders. The following are aspects of the value.

\subsubsection{Independent peer review}
Peer review is commonly used in quality regimes in higher education and in periodic review processes. In most cases the HEI being reviewed chooses the peer. As part of a BCS (or other professional body) a HEI has less choice regarding who completes a review and hence such a review is arguably more truly independent. BCS review panels contain a minimum of two experienced assessors who have strong awareness of the discipline norms across the UK sector.

\subsubsection{Internationally agreed standards}
The BCS accredits to internationally agreed standards, which evidence the global parity of the degree programmes accredited.  This assists in the global mobility for the graduates of accredited programmes.

\subsubsection{Ensuring curricula relevance}
With the agreement of the sector (normally reached through the Council of Professors and Heads of Computing (CPHC)), it is possible to agree and promote curricula change. The most recent example of this is the inclusion of cybersecurity in all accredited degrees \cite{Cricketal2019}. BCS is in the process of mandating the inclusion of security in all accredited degree programmes to a standard agreed between professional bodies, industry and government. This is a work in progress \cite{Irons2016}. BCS has been requiring coverage since 2015 \cite[p.~17--18]{BCS2018a} with the result that all accredited universities must be compliant by 2020 (due to the five-year cycle). In the United States of America, the Association of Computing Machinery (ACM) has equally had cybersecurity (IAS: ``Information Assurance and Security'') in the curriculum since 2013~\cite{ACM2013a}, but it is not the accrediting body. The Accreditation Board for Engineering and Technology (ABET) is, and is requiring IAS with effect from the 2019-20 cycle (self-study reports due 1 July 2019): more precisely \cite[Table 3]{Oudshoornetal2018a} ``{\emph{The computing topics must include: \dots{} Principles and practices for secure computing\dots}}''. This will mean that  all accredited universities should be compliant by 2025 (due to their six-year cycle).

It is interesting to note this years Royal Institution Christmas lecture considers ethical questions related to computing and mathematics \cite{RoyalInstitution2019}. This is a curricula element that has been mandated by BCS accreditation in the form of legal, social, ethical and professional issues for a number of years \cite{Brooke2018}.

\subsubsection{Disseminating good practice}
BCS accreditation panels have been identifying aspects of good practice for wider dissemination for a number of years. By commending good practices and promoting these to the wider community. Since 2016, the process has been refined to more clearly signpost commendations. BCS is currently in the process of refining the promotion of good practice.

One short term initiative, from autumn 2019, visited HEIs will be offered the opportunity to showcase an aspect of their provision or practice. The intention is to generate further good practice examples that a visiting panel can potentially commend.

\subsubsection{Industry relevance}
All BCS accreditation panels include an industrial assessor whose role is to ensure that programmes appropriately meet the needs of industry. Evidence that a visited HEIs' mechanisms for engaging with industry are supporting graduates to evidence they are achieving the expected exit standards are sought from a variety of sources including the curricula studied; the assessments and examinations sat; and the engagement mechanisms themselves.

\subsubsection{Maintaining standards}
BCS refuses accreditation for programmes that do not meet the expected standards or do not comply with the curricula expectations of the accreditation sought.

\subsection{Driving improvement}
A number of data sources have been employed to help drive the improvement efforts. BCS Secretariat continually evaluate via the use of opinion surveys and informal conversations. Views are gathered related to briefing sessions, pre-visit communications and visits. This information is fed into the ongoing reviews. These issues are discussed, where required at an appropriate committee meeting and used to identify any enhancement projects. Following a visit BCS panels engage in peer review. The feedback gained from assessor peer review is explored and opportunities for enhancement are discussed on an annual basis at a committee meeting.

\subsection{Reducing bureaucracy and enhancing practices}
A review of BCS Accreditation practice has been taking place since 2015 with the intention of adopting an enhancement oriented agenda; reducing the amount of bespoke documentation that is required; using technology to assist the process, whilst enforcing the international standards.

Working through the process from a visited HEIs prospective the review results in the following changes.

\subsubsection{HEI briefing}
For a number of years all visited HEIs have been encouraged to attend a briefing to explore the BCS requirements with a focus upon any changes since the last visit and common challenges that present themselves as part of visits. Attendance was understandably mixed, partly due to travel. Since 2017 all Briefings have taken place by video conference. Feedback upon this approach has been very positive.

\subsubsection{The application itself}
From a paper and memory stick / CD submission before 2015, BCS has moved to a fully electronic submission. BCS is flexible in how the information is provided, a range of secure cloud based file sharing systems have been employed, submission via the use of Virtual Learning Environment has been completed and a minority of institutions have opted to create a website related to the submission. Most of the application consists of an evidence base which HEIs will already have. This is supplemented by a summary of the provision and each programme in which the department are welcome to reference existing resources. HEIs are still required to provide a mapping of where the BCS requirements are taught and assessed within each programme. This is required to evidence the where accreditation specific requirements are met in a particular programme.

\subsubsection{Problematic areas}
It is now normal practice for a BCS panel to communicate likely areas of discussion to a visited HEI prior the visit. There are some occasions when this is not possible, for example if a HEIs application was late or if other issues emerge as part of the visit.   This has resulted in discussions tending to become more collegiate.

\subsection{Promoting the value of accreditation}
The value proposition discussed in this document is a definition of the value of accreditation from those who are intimately involved. BCS is in the process of initiating a conversation to explore the value proposition of accreditation from the point of view of other stakeholders.

\section {Where does it fit?	}
%A short description of your teaching context. You may, for instance, include a description of intake, class size, curriculum sequence; anything that's necessary for others to understand your situation. How do things work at your institution?

In the United Kingdom the most common form of accreditation in the Computer Science and related discipline areas is by professional bodies.  The BCS, The Chartered Institute for IT (BCS) and the Institution of Engineering and Technology (IET) are the main bodies involved. The accreditation provided by these institutes are underpinned by international Initiatives such as the Washington Accord \cite{Washington2019} and Seoul Accord \cite{Seoul2019}. These memoranda support the internationalising of the curriculum and promote consistency and parity in Computer Science education globally. These professional bodies are also registered charities and hence have responsibilities for public good which extends beyond short term job market needs \cite{Stensaker2006}, \cite{Mutereko2017}. This is reflected in IETs and BCSs long history of expecting coverage environmental factors within the programmes they accredit and BCSs expectation of significant coverage of Legal, Ethical, Social and Professional Issues \cite{Brooke2018}.

As part of accreditation IET and BCS broadly check two things \cite{BCS2018a, IET2019}.

\begin{enumerate}
    \item Is the quality of experience provided by a HEI appropriate for accreditation?
    A number of sources of information are used to determine this and a number of metrics are explored, including graduate employability.  This is supplemented by other evidence of the quality of the provision, for example external examiners reports, the most recent subject review, annual review information, evidence of employer involvement, linkage with research. Together this evidences that a HEI is supporting their students in an appropriate manner. Accreditation will be refused for HEIs that do not offer appropriate support (including for example poor employability of graduates from a programme).  
    \item Are the exit standards of a programme consistent with the learning outcomes expected for the accreditation sought? The exit standards conform with one of the international memorandum (Washington or Seoul Accord or both). 
\end{enumerate}

\section {Does it work?}	H
%how do you know? Give some evidence of effectiveness in context.

HEIs choose which programmes are submitted or not in an accreditation application. As such only a sample of provision may be considered.

However a proxy for success is the anonymous feedback BCS obtains post visit. Some key aspects of this feedback are shown in the following table.

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}
%\begin{table}[]
\begin{table}[h!]
  \caption{Selected results from accreditation visit feedback survey}
  \label{table:1}
%\resizebox{\textwidth}{!}{%
%\begin{tabular}{lllll}
\begin{tabular}{ | p{5cm}|p{.75cm}|p{.75cm} |p{.75cm} |}
\hline
 & 2017 & 2018 & 2019   \\ \hline
Responses & 12 & 14 & 16   \\
\hline
Attended pre-visit briefing & 6 & 14 & 16   \\
\hline
The visit felt to be worthwhile: & & &  \\
Strongly Agree &92\% & 93\% &88\% \\
Agree &0\% & 7\% &0\% \\
Tend to Agree &0\% & 0\% &6\% \\
Strongly Disagree &8\% & 0\% &0\% \\
No reply &0\% & 0\% &6\% \\
\hline
Overall how satisfied  would you say you were with the visit? & & &  \\
Very Satisfied &83\% & 86\% &94\% \\
Quite Satisfied &8\% & 14\% &6\% \\
Not Satisfied at all &8\% & 0\% &0\% \\
\hline

\end{tabular}%
%}
\end{table}

Overall the results are positive, indicating the process is generally value and shows the increased attendance of the virtual briefing over the physical one.

\section {Who else has done this?}	
%Where did you get the idea from? (If from published reports, please include references). How did you find out about it? Was it easy/hard to adopt? What did you change?

In addition IET and BCS, accreditation in the broad computing area in the UK is also being performed by a few different agencies.

\subsubsection{Not for Profit Organisations}
Tech Partnership Degrees is a Not For Profit Organisation that provides endorsements to Higher Education programmes with specific curricula elements aimed at job market requirements \cite{TP2019}. Tech Partnership Degrees have a specialist scope, endorsing programmes in the area of IT Management for Business and Software Engineering for Business.  Tech Partnership Degrees currently endorse 14 IT Management for Business Programmes and 5 Software Engineering for Business Programmes. As such Tech Partnership Degrees currently have limited impact upon more general Computer Science education. There is little published regarding the effectiveness or otherwise of this approach.

The Institute of Coding (IoC) is a not for profit organisation that intends to enhance how Digital Skills are developed in Higher Education in the UK \cite{Davenportetal2019a}. Like the Tech Partnership focus is upon job market requirements. A micro-credentialing approach appears to being taken. It still is in the early stages of adoption. The BCS is actively working with IoC and intends to collaborate with the IoC in any initiatives of mutual benefit.

\subsubsection{National Cyber Security Centre (NCSC)}
The NCSC is a UK Government organisation tasked with enhancing the cybersecurity of the UK. The NCSC publishes and accredits to a number of cybersecurity standards \cite{NCSC2018a}. These standard are linked to the Association of Computer Machinary (ACM) recommendations for curricula. To date the major focus of NSSC accreditation has been upon Masters degrees specializing in cybersecurity. More recently the NCSC has also began accrediting integrative masters programmes, undergraduate degrees in cybersecurity and Computer Science degrees with a significant cybersecurity focus \cite{NCSC2018b}.

Currently accredited are 15 cybersecurity MSc programmes with a further 11 provisional accredited, 3 integrated masters cybersecurity programmes and 1 cybersecurity Degree Programme with a further 2 provisionally accredited. Hence the extent of accreditation is currently reasonably limited in terms of reach to Computer Science programmes. This appears to be a positive initiative that will hopefully further develop over time. Again little has yet been published regarding the effectiveness of this initiative.

\subsubsection{The wider Computer Science education community}
The Computer Science discipline has responded actively to the challenges presented by Shadbolt with CPHC operating four working groups \cite{cphc_2016},  promotion of further engagement between industry and academia by initiatives such as the Royal Academy of Engineering Visiting Industrial Professor Scheme \cite{royal}

\section {What will you do next?}	
%Will you vary this, or develop it further?

BCS reviews its accreditation processes on a regular basis. The current enhancement projects that are either in progress or planed for the future are as follows.

\subsection{Internationalization}
BCS accreditation primarily takes place in the UK, however there are a small but growing number of institutions outside the UK that BCS accredits or who are seeking accreditation from BCS. Under the terms of the Washington and Seoul Accords, BCS does not engage in accreditation in the jurisdictions of Accord signatories without first consulting with the related local professional body.  On occasions BCS guidelines employ English / Welsh higher education terms without indicating that local equivalents are equally acceptable. The next set of guidelines will address this short coming.

\subsection {Engineering Council and Accreditation of Higher Education Programmes (AHEP) Version 4}
In the British jurisdiction of the Washington accord, the approach adopted is unusual in that the Engineering Council extends the license to accredit to a large number of professional bodies \cite{EC2019}. This is not the practice adopted in other jurisdictions, for example in the United States of America ABET is the sole body. A recent audit by the Washington Accord has highlighted divergence in practice in some areas with respect to Chartered Engineer Accreditation. The Engineering Council is in the process of seeking further consistency. This has lead to the new rules with respect of Compensation and Condonement for Eng accreditation \cite{EC2018}. Further changes are embedded in AHEP 4. The BCS requirements have been updated to reflect the expectations of Compensation and Condonement. Once AHEP 4 is finalised further work will be required to address these updates. Among other proposals, AHEP 4 intends to extend Chartered Engineering accreditation to include diversity and widening participation data (of staff and students) as part of the metrics that assess the quality of provision. The Computer Science discipline could clearly do better in this regard, so this should be a positive inclusion.

\subsection{Support for HEIs seeking accreditation}
As part of its operation BCS continually reviews and monitors the support process it employs for HEIs seeking accreditation and hence will continue to enhance processes in this area.

\subsection{Agility of processes and changing external environment}
There are considerable external pressures placed upon HEIs, by for example the Research Excellent Framework (REF) and the Teaching Excellence Framework(TEF), which could reduce the priority placed upon accreditation as a mechanism for supporting enhancement. Care is taken in visit reports to ensure positive aspects of provision are emphasised. This is in part intend to provide evidence that could be used to support future TEF related submissions. Discussions regarding the agility of processes are incorporated into two annual committee meetings to identify short, medium and long term projects.

\subsection {Further role for employers in accreditation}
As indicated previously an industrial assessor is a critical part of every visit panel. There are also interesting developments being lead by the IoC. Specifically, the IoC is proposing micro-credentialing as an extension to accreditation. This could potentially augment the current recognised pathways to professional accreditation by providing a more fine grained alternatives that could be useful to some employers or employees who wish to evidence their achievements in an accredited manner. The details of this proposal is not finalised, however the BCS continues to explore possibilities with the IoC with a view to supporting any initiative of mutual benefit. One of the challenges in this work is the lack of an agreed standard (in the way there is for the Seoul and Washington Accords) for micro-credentialing.

\subsection {Exploration of the value of accreditation to all stakeholders}
As discussed earlier,  BCS AAC is in the process of initiating a conversation to explore the value proposition of accreditation from other stakeholders points of view.

\subsection {Improve communications}
There is a strong and vibrant community of academics and industrialists who actively participate in accreditation. However, there are many involved with whom communication could benefit from being more active. A number of initiatives are in process to enhance these communications. For HEIs there are ongoing discussion within BCS regarding establishing a periodic bulletin which could be used to share updates of accreditation practice / procedure, highlight HEIs having their first visit, sharing good practice examples, promote assessor recruitment and so on. Assessor Peer Review has indicated more regular communication with assessors could be of benefit. Virtual link up sessions for assessors are proposed. The BCS are in the process of contacting assessors to determine likely interest and thoughts related to relevant topics for such link ups.

\subsection {Mechanisms for sharing good practice}
Whilst BCS has been identifying aspects of good practice as part of accreditation visits for a few years. They mechanism (publishing on the BCS website) does not appear to have been very effective.  There are conversations ongoing regarding how this can be improved.

\section{Why are you telling us this?}	
%What is interesting, or useful, about this to someone else?

One of the recommendations of Shadbolt is the value of accreditation should be more clearly communicated to stakeholders. This paper is part of a set of initiatives to achieve precisely that.

Much has changed in BCS accreditation in the last few years. For many academics their only experience of BCS accreditation is the quinquennial BCS accreditation visit. Accreditation by the BCS is performed by panels of BCS assessors. An assessor may be an industrialist or an academic. The size of a panel varies depending upon the number of programmes a visited HEI is putting forward for consideration. A panel will always include an industrial assessor and two or more academic assessors.  At most HEIs will have two academic assessors.  There is a large pool of academic assessors and as such not all assessors will complete an accreditation visit every year. Not all HEIs have an academic assessor. Hence a significant set of the Computer Science education community are not informed regarding the evolution of accreditation practice.

As such the intention of the paper is share to the wider community the developments made, the future enhancement aspirations and the forthcoming opportunities to engage in further conversations.
